Parties, Models, Mobsters\newline {\fontsize{11}{15}\selectfont A New Implementation of Model-Based Recursive Partitioning in R}

### 2014-03-09 pkg/partykit/inst/NUS-2014/modelparty.Rnw
merged slides from JKU and Psychoco -> still needs a lot of work
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@465 255f4249-2263-4417-b1ea-8498779e27a7

\graphicspath{{images/}}

@@

\subsection{Model-based recursive partitioning}

\begin{frame}
\frametitle{Model-based recursive partitioning}

\textbf{Models:} Estimation of parametric models with
observations $y_i$ (and regressors $x_i$),
parameter vector $\theta$, and additive objective function $\Psi$.
\[ \widehat \theta \quad = \quad \argmin_{\theta}
   \sum_i \Psi(y_i, x_i, \theta). \]

@@

\textbf{Recursive partitioning}:

\begin{enumerate}
  \item Fit the model in the current subsample.
  \item Assess the stability of $\theta$ across each partitioning variable $z_j$.
  \item Split sample along the $z_{j^*}$ with strongest association:
        Choose breakpoint with highest improvement of the model fit.
  \item Repeat steps 1--3 recursively in the subsamples
        until some stopping criterion is met.
\end{enumerate}

@@

\begin{frame}
\frametitle{Model-based recursive partitioning}

\textbf{Parameter instability tests:}
\begin{itemize}
  \item Based on empirical estimating functions (or score/gradient contributions):
    $\Psi'(y_i, x_i, \hat \theta)$.
  \item Under parameter stability: $\Psi'$ fluctuates randomly around its expectation zero.
  \item Under parameter instability: Systematic departures from zero in subsamples.
  \item Hence fluctuation can be captured across numeric partitioning variables
    or within levels of categorical partitioning variables.
  \item Bonferroni correction for testing across multiple partitioning variables.
\end{itemize}

>>>>

\graphicspath{{../Psychoco-2014/images/}}

@@

\subsection{Motivation}

\begin{frame}
\frametitle{Motivation: Trees}

Breiman (2001, \textit{Statistical Science}) distinguishes two cultures
of statistical modeling.

\begin{itemize}
  \item \textbf{Data models:} Stochastic models, typically parametric.\\
        $\rightarrow$ Classical strategy in statistics.
	Regression models are still the workhorse for many empirical analyses.
  \item \textbf{Algorithmic models:} Flexible models, data-generating process unknown.
        $\rightarrow$ Still few applications in many fields, e.g., social sciences or economics.
\end{itemize}

\textbf{Examples:} Recursive partitioning, decision trees.
Class of flexible methods for classification and regression.

\begin{frame}
\frametitle{Motivation: Trees}

\textbf{Formally}: Modeling of dependent variable $Y$ by ``learning''
a recursive partition w.r.t\ explanatory variables $Z_1, \dots, Z_l$.

@@

\textbf{Key features}:

\begin{enumerate}
  \item Predictive power in nonlinear regression relationships.
  \item Interpretability (enhanced by visualization), i.e., no ``black box'' methods.
\end{enumerate}

\textbf{Examples}: CART and C4.5 in statistical and machine learning, respectively.

@@

\begin{frame}
\frametitle{Motivation: Leaves}

\textbf{Typically:} Simple models for univeriate $Y$, e.g., mean.

\textbf{Idea:} More complex models for multivariate $Y$, e.g., multivariate
normal model, regression models, etc.

\textbf{Here:} Synthesis of parametric data models and algorithmic tree models.

\textbf{Goal:} Fitting local models by partitioning of the sample space.

\subsection{Recursive partitioning}

\begin{frame}
\frametitle{Recursive partitioning}

\textbf{Base algorithm}:

\begin{enumerate}
  \item Fit model for $Y$.
  \item Assess association of $Y$ and each $Z_j$.
  \item Split sample along the $Z_{j^*}$ with strongest association:
        Choose breakpoint with highest improvement of the model fit.
  \item Repeat steps 1--3 recursively in the subsamples
        until some stopping criterion is met.
\end{enumerate}

%% \medskip
%% 
%% \textbf{Generally:} Tree algorithms differ w.r.t.\ choice of model (1),
%% association measure~(2), split strategy (3) and stopping criterion (4).

\textbf{Here:} Segmentation~(3) of parametric models~(1) with 
additive objective function using parameter instability tests~(2)
and associated statistical significance~(4).

\subsection{Model estimation}

\begin{frame}
\frametitle{1.~Model estimation}

\textbf{Models:} $\mathcal{M}(Y, \theta)$ with (potentially) multivariate
observations $Y \in \mathcal{Y}$ and $k$-dimensional parameter vector
$\theta \in \Theta$.

\textbf{Parameter estimation:} $\widehat \theta$ by optimization of
objective function $\Psi(Y, \theta)$ for $n$ observations $Y_i$ ($i = 1, \dots, n$):

\[ \widehat \theta \quad = \quad \argmin_{\theta \in \Theta}
   \sum_{i = 1}^n \Psi(Y_i, \theta). \]

\textbf{Special cases:} Maximum likelihood (ML), weighted and ordinary
least squares (OLS and WLS), quasi-ML, and other M-estimators.

\begin{frame}
\frametitle{1.~Model estimation}

\textbf{Estimating function:} $\widehat{\theta}$ can also be defined in terms of
  \[ \sum_{i = 1}^n \psi(Y_i, \widehat{\theta}) = 0, \]
where $\psi(Y, \theta) = {\partial \Psi(Y, \theta)}/{\partial \theta}$.

\textbf{Central limit theorem:} If there is a true parameter $\theta_0$ and given
certain weak regularity conditions:
  \[ \sqrt{n} (\widehat{\theta} - \theta_0) \quad \darrow \quad \mathcal{N}(0, V(\theta_0)), \]
where $V(\theta_0) = \{A(\theta_0)\}^{-1} B(\theta_0) \{A(\theta_0)\}^{-1}$.
$A$ and $B$ are the expectation of the derivative of $\psi$ and
the variance of $\psi$, respectively.

\begin{frame}
\frametitle{1.~Model estimation}

\textbf{Idea:}
In many situations, a single global model 
$\mathcal{M}(Y, \theta)$ that fits \textbf{all} $n$~observations cannot be found.
But it might be possible to find a partition w.r.t.\ the variables
$Z = (Z_1, \dots, Z_l)$ so that a well-fitting model can be found locally
in each cell of the partition.

\textbf{Tool:} Assess parameter instability w.r.t\ to partitioning variables
$Z_j \in \mathcal{Z}_j \; (j = 1, \dots, l)$.

\subsection{Parameter instability tests}

\begin{frame}
\frametitle{2.~Tests for parameter instability}

Generalized M-fluctuation tests
capture instabilities in $\widehat \theta$ for an ordering w.r.t\ $Z_j$.

\textbf{Basis:} Empirical fluctuation process of cumulative deviations
w.r.t.\ to an ordering $\sigma(Z_{ij})$.

\[
W_j(t, \widehat \theta)
  \quad = \quad {\widehat B}^{-1/2} n^{-1/2}
                \sum_{i = 1}^{\lfloor nt \rfloor}
		\psi(Y_{\sigma(Z_{ij})}, \widehat \theta)
\qquad (0 \le t \le 1)
\]

\textbf{Functional central limit theorem:} Under parameter stability
$W_j(\cdot) \, \darrow \, W^0(\cdot)$, where
$W^0$ is a $k$-dimensional Brownian bridge.

\begin{frame}
\frametitle{2.~Tests for parameter instability}

\setkeys{Gin}{width=1.08\textwidth}
\hspace*{-0.7cm}%
<<fig=TRUE, height=5, width=10, echo=FALSE, results=hide>>=
## artificial data
set.seed(42)
a <- rnorm(41, mean = 0, sd = 0.12)
b <- rnorm(41, mean = 1, sd = 0.12)
b[40] <- b[40]-0.1
b[41] <- b[41]-0.2
a[1] <- a[1]+0.2
a[2] <- a[2]+0.1
x <- c(b, a)

par(mar = c(5, 5, 3, 1), mfrow = c(1, 2))

## data, model, and residuals
plot(0:81, x, xlab = expression(Z[j]), ylab = "Y", ylim = c(-2, 3),
  type = "l", xlim = c(1, 80), axes=FALSE)
axis(1,at=seq(0,80,20),labels=seq(2004,2012,2),cex.axis=1.2, cex.lab=1.2)
axis(2,at=seq(-2,3,1),labels=seq(200,1200,200),cex.axis=1.2, cex.lab=1.2)
box()
abline(h = mean(x), lty = 3)
for(i in c(5, 15, 25, 35)) {
  segments(x0 = i, x1 = i, y0 = x[i+1], y1 = mean(x), lty = 2)
  i <- i + 40
  segments(x0 = i, x1 = i, y0 = mean(x), y1 = x[i+1], lty = 2)
}

## CUSUM graph
x_efp <- c(0, cumsum(x - mean(x)))/sqrt(length(x) * var(x))
plot(-1:81, x_efp, xlab = expression(Z[j]), ylab = "Fluctuation Process", 
  type = "l", xlim = c(1, 80), ylim = c(-5, 5), axes=FALSE)
axis(1,at=seq(0,80,20),labels=seq(2004,2012,2),cex.axis=1.2, cex.lab=1.2)
axis(2, cex.axis=1.2, cex.lab=1.2)
box()
abline(h = 0, lty = 3)
for(i in seq(5, 75, by = 10)) segments(x0 = i, x1 = i, y0 = 0, y1 = x_efp[i+2], lty = 2)
@

\begin{frame}
\frametitle{2.~Tests for parameter instability}

\textbf{Test statistics:} Scalar functional $\lambda(W_j)$
that captures deviations from zero.

\textbf{Null distribution:} Asymptotic distribution of $\lambda(W^0)$.

\textbf{Special cases:} Class of test encompasses many well-known
tests for different classes of models.
Certain functionals~$\lambda$ are particularly intuitive for numeric and
categorical $Z_j$, respectively.

\textbf{Advantage:} Model $\mathcal{M}(Y, \widehat \theta)$ just has to be
estimated once. Empirical estimating functions $\psi(Y_i, \widehat \theta)$
just have to be re-ordered and aggregated for each $Z_j$.

\begin{frame}
\frametitle{2.~Tests for parameter instability}

\textbf{Splitting numeric variables:} Assess instability using
sup$\mathit{LM}$ statistics.

\[
\lambda_{\mbox{sup}\mathit{LM}} (W_j) \quad = \quad
\max_{i = \ui, \dots, \oi}  \, \left(\frac{i}{n} \cdot \frac{n-i}{n} \right)^{-1}
\left| \left| W_j \left( \frac{i}{n} \right) \right| \right|^2_2.
\]

\textbf{Interpretation:} Maximization of single shift $\mathit{LM}$ statistics
for all conceivable breakpoints in $[\ui, \oi]$.

\textbf{Limiting distribution:} Supremum of a squared, $k$-dimensional tied-down
Bessel process.

\begin{frame}
\frametitle{2.~Tests for parameter instability}

\textbf{Splitting categorical variables:} Assess instability using
$\chi^2$ statistics.

\[ \lambda_{\chi^2} (W_j) \quad = \quad
\sum_{c = 1}^C 
\frac{n}{|I_c|}  \left| \left| \Delta_{I_c} W_j \left( \frac{i}{n} \right) \right| \right|^2_2
\]

\textbf{Feature:} Invariant for re-ordering of the $C$ categories and the
observations within each category.

\textbf{Interpretation:} Captures instability for split-up into $C$ categories.

\textbf{Limiting distribution:} $\chi^2$ with $k \cdot (C-1)$ degrees of freedom.

\begin{frame}
\frametitle{2.~Tests for parameter instability}

\textbf{Splitting ordinal variables:} FIXME

like categorical

like numeric but invariant to reordering within ties

\subsection{Segmentation}

\begin{frame}
\frametitle{3.~Segmentation}

\textbf{Goal:} Split model into $b = 1, \dots, B$ segments along the
partitioning variable $Z_j$ associated with the highest parameter instability.
Local optimization of

\[ \sum_b \sum_{i \in I_b} \Psi(Y_i, \theta_b). \]

$B = 2$: Exhaustive search of order $O(n)$.

$B > 2$: Exhaustive search is of order $O(n^{B-1})$, but can be replaced
by dynamic programming of order $O(n^2)$. Different methods (e.g., information
criteria) can choose $B$ adaptively.

\textbf{Here:} Binary partitioning.

\subsection{Pruning}

\begin{frame}
\frametitle{4.~Pruning}

\textbf{Pruning:} Avoid overfitting.

\textbf{Pre-pruning:} Internal stopping criterium. Stop splitting when
there is no significant parameter instability based on
Bonferroni-corrected $p$~values of the fluctuation tests.

\textbf{Post-pruning:} Grow large tree and prune splits that do not
improve the model fit based on information criteria (e.g., AIC or BIC).
